Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-15 opencl/papers/programming-model.tex
cleaning up after Craig....

Unlike OpenMP, we require limited use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:
\begin{itemize}
   \item !$OFP CONTIGUOUS: specifies that an dummy array variable is
   contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.
   \item !$OFP KERNEL: specifies that a pure subroutine can be
     transformed to an OpenCL kernel.
\end{itemize}

Three new procedures (in addition to the intrinsic shift function) are
defined in Fortran that are used in array-syntax operations.  Each
procedure takes a integer array halo argument that specifies the
number of ghost cells on either side of a region, for each dimension.
For example {\tt halo = [left, right, down, up]} specifies a halo for
a two-dimensional region.  These functions are:
\begin{itemize}
   \item transfer_halo(array, halo): an impure subroutine that
     exchanges halo regions between nodes using MPI or Fortran
     coarrays. Array is the memory stored on the node representing the
     node-local portion of a virtual global array. Not used in this work.
   \item interior(array, halo): a pure function that returns a copy of
     the interior portion of the array specified by halo. Array is
     local to a node.
   \item region(array, halo): a pure function that returns a copy of
     the portion of the array specified by halo.  Array is local to a node.
\end{itemize}

@@

A kernel procedure (specified by the !$OFP KERNEL directive) must be
pure and all array variables must be declared as contiguous.  A kernel
procedure may not call other procedures except for limited intrisic
functions (primarily math), user-defined elemental function, and the
{\tt interior} and {\tt region} functions.

>>>>

Unlike OpenMP, we require limited use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:
\begin{itemize}
   \item {\tt \!\$OFP CONTIGUOUS}: specifies that an dummy array variable is
   contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.
   \item {\tt \!\$OFP KERNEL}: specifies that a pure subroutine can be
     transformed to an OpenCL kernel.
\end{itemize}

Three new procedures (in addition to the intrinsic shift function) are
defined in Fortran that are used in array-syntax operations.  Each
procedure takes a integer array halo argument that specifies the
number of ghost cells on either side of a region, for each dimension.
For example {\tt halo = [left, right, down, up]} specifies a halo for
a two-dimensional region.  These functions are:
\begin{itemize}
   \item {\tt transfer\_halo(array, halo)}: an impure subroutine that
     exchanges halo regions between nodes using MPI or Fortran
     coarrays. Array is the memory stored on the node representing the
     node-local portion of a virtual global array. Not used in this work.
   \item {\tt interior(array, halo)}: a pure function that returns a copy of
     the interior portion of the array specified by halo. Array is
     local to a node.
   \item {\tt region(array, halo)}: a pure function that returns a copy of
     the portion of the array specified by halo.  Array is local to a node.
\end{itemize}

@@

A kernel procedure (specified by the {\tt \!\$OFP KERNEL} directive) must be
pure and all array variables must be declared as contiguous.  A kernel
procedure may not call other procedures except for limited intrisic
functions (primarily math), user-defined elemental function, and the
{\tt interior} and {\tt region} functions.

### 2010-12-15 opencl/papers/source-to-source.tex
cleaning up after Craig....

Fortran kernel procedures have direct correspondence with OpenCL equivalents.
For example, {\tt wave_advance} is transformed as {\tt __kernel void
  wave_advance(__ global float * H, ..., float dt);}.  The global state
variables have local equivalents, e.g., {\tt __local float
  H_local[MAX_LOCAL_SIZE]}, declared with the appropriate size and are copied
to local memory by the compiler with inlined library functions.  The flux
array temporaries (region variables) are declared similarly but initialized to
zero with inlined library functions.  Interior variables are simple scalars,
e.g., {\tt float iH;}.  Region variables cannot be scalar objects because
regions are shifted and thus \emph{shared} by threads within a thread group.

@@

Region variables are more complicated because they are arrays.
\begin{verbatim}
   Hx[l] = 0.5 * ( region(H_local, face_lt) + ... );
\end{verbatim}
where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
  LX = get_local_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
  get_local_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
  get_local_size(0)} is the size of the thread group in the $x$ dimension, and
the {\tt get_local_id} and {\tt get_local_size} functions are defined by the
OpenCL compiler.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{cl-performance.pdf}
\caption{Performance comparison for varying array size.}
\label{fig:cl-performance}
\end{figure}

>>>>

Fortran kernel procedures have direct correspondence with OpenCL equivalents.
For example, {\tt wave\_advance} is transformed as {\tt \_\_kernel void
  wave\_advance(\_\_ global float * H, ..., float dt);}.  The global state
variables have local equivalents, e.g., {\tt \_\_local float
  H\_local[MAX\_LOCAL\_SIZE]}, declared with the appropriate size and are copied
to local memory by the compiler with inlined library functions.  The flux
array temporaries (region variables) are declared similarly but initialized to
zero with inlined library functions.  Interior variables are simple scalars,
e.g., {\tt float iH;}.  Region variables cannot be scalar objects because
regions are shifted and thus \emph{shared} by threads within a thread group.

@@

Region variables are more complicated because they are arrays.
\begin{verbatim}
   Hx[l] = 0.5 * ( region(H_local, face_lt) + ... );
\end{verbatim}
where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
  LX = get\_local\_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
  get\_local\_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
  get\_local\_size(0)} is the size of the thread group in the $x$ dimension, and
the {\tt get\_local\_id} and {\tt get\_local\_size} functions are defined by the
OpenCL compiler.

@@

%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{cl-performance.pdf}
%\caption{Performance comparison for varying array size.}
%\label{fig:cl-performance}
%\end{figure}
